---
title: "META-PRISM: Define colors common to all plots"
author: "Yoann Pradat"
date: "`r format(Sys.time(), '%d %B, %Y')`"
link-citations: true
output:
  bookdown::html_document2:
    theme: cosmo
    toc: true
    toc_float: true
    code_folding: show
  rmdformats::material:
    highlight: kate
---

<style>
body {
text-align: justify}
</style>

```{r set-options, echo=FALSE, cache=FALSE}
options(width=120)
```

# Technical setup 

## Libraries

R libraries.

```{r R libraries, message=F}
# using python in R
library(reticulate)

# save
library(xlsx)

# in-house packages
library(rprism)
library(rprismtools)
```

Python libraries.

```{python python libraries}
import pandas as pd
import numpy as np
from pyprism.data import load_cln, load_bio

import matplotlib
import matplotlib.cm as cm
```

## Functions

R functions

```{r R functions}
add_colors_to_table <- function(colors, name, dfs=NULL){
  if (is.null(dfs)){
    dfs <- list()
  }
  
  if (name %in% names(dfs)){
    n <- nrow(dfs[[name]])
    df <- data.frame(Name=names(colors), Color=unlist(colors), row.names=seq(n+1, n+length(colors)))
    dfs[[name]] <- rbind(dfs[[name]], df)
  } else {
    df <- data.frame(Name=names(colors), Color=unlist(colors), row.names=seq(1, length(colors)))
    dfs[[name]] <- df
  }

  dfs
}
```


Python functions

```{python python functions}
def get_labels2colors(labels_unique, sort=False, cmap_name=None, palette=None, alpha=None, colors_avoid=[]):
    labels2colors = dict()

    if palette is None and cmap_name is None:
        raise ValueError("Specify either palette or cmap_name")
    elif palette is None:
        cmap = cm.get_cmap(cmap_name)
        palette = [cmap(i) for i in range(cmap.N)]

    if sort:
        labels_unique = sorted(labels_unique)

    i = 0
    for label in labels_unique:
        color = palette[i % len(palette)]
        while color in colors_avoid:
            i = i+1
            color = palette[i % len(palette)]
        color = matplotlib.colors.to_hex(color)
        if alpha is not None:
            op = re.sub("0x", "", hex(round(alpha*255)))
            if len(op)==1:
                op = "0" + op
            color = color + op
        labels2colors[label] = color.upper()
        i = i+1

    return labels2colors
```

```{r init colors list}
dfs_colors <- list()
```

## Parameters

```{r parameters}
folder_results <- "../../../../results/data_overview/colors"
if (!dir.exists(folder_results)){
  dir.create(folder_results, showWarnings=F, recursive=T)
}
```


# Global colors

Let us first define colors that will serve in the graphics parameters of multiple analyses. These include the colors for
representing each of the pancancer cohorts (MET500, PRISM, TCGA), the colors for representing tumor types, the colors
for representing biopsy sites, etc. Once the colors are defined, they are depicted in this notebook and saved in a
common excel worksheet.

Here are the colors used for

```{r colors cohorts, fig.width=6, fig.height=1, fig.align="center"}
cohorts2colors <- list(prism="#E5383B", met500="#90BE6D", tcga="#577590")
cohorts2colors <- cohorts2colors[sort(names(cohorts2colors))]
dfs_colors <- add_colors_to_table(cohorts2colors, name="Global", dfs=dfs_colors)
rect_plot_colors(line=1, colors=cohorts2colors)
```

```{r colors bicohorts, fig.width=6, fig.height=1.5, fig.align="center"}
bicohorts2colors <- list(prism_clear="#F6A38E", prism_dark="#D03711",
                         met500_clear="#74C69D", met500_dark="#1B4332",
                         tcga_clear="#48cae4", tcga_dark="#023E8A")
bicohorts2colors <- bicohorts2colors[sort(names(bicohorts2colors))]
dfs_colors <- add_colors_to_table(bicohorts2colors, name="Global", dfs=dfs_colors)
rect_plot_colors(line=3, colors=bicohorts2colors)
```

# Clinical fields colors

```{python load clinical}
cohorts = ["prism", "met500", "tcga"]
dfs_cln = {cohort: load_cln(study=cohort, mode="in_design") for cohort in cohorts}

# fill na
cols_fillna = {"Primary_Site", "Project_TCGA", "Project_TCGA_More", "Histological_Type"}

for name, df in dfs_cln.items():
    cols_df = set(df.columns)
    for col in cols_fillna.intersection(cols_df):
        df[col] = df[col].fillna("N/A").astype(str)
```

```{python clinical colors}
def get_dict_mode_xy(df, x, y):
    s_mode = df.dropna(subset=[x,y]).groupby(x)[y].agg(lambda x: pd.Series.mode(x).iat[0])
    df_mode = s_mode.to_frame(y).reset_index()
    return {r[x]:r[y] for _, r in df_mode.iterrows()}

# Project_TCGA colors
PT_unique = set().union(*[set(df["Project_TCGA"]) for df in dfs_cln.values()])
PT2colors = get_labels2colors(labels_unique=PT_unique, sort=True, cmap_name="tab20")
PT2colors["LUSC"] = "#DBDB8D"
PT2colors["Unknown_Primary"] = "#D3D3D3"

# Project_TCGA_More colors
PTM_unique = set().union(*[set(df["Project_TCGA_More"]) for cohort, df in dfs_cln.items() if cohort!="tcga"])
PTM_unique.remove("MISC - Not_TCGA")

PTM2colors_ntcga = {"HNAC - Not_TCGA": PT2colors["HNSC"], "LUNE - Not_TCGA": PT2colors["LUSC"],
                    "SARC - Not_TCGA": PT2colors["SARC"], "BLCA - Not_TCGA": PT2colors["BLCA"],
                    "THCA - Not_TCGA": PT2colors["THCA"], "MISC - Not_TCGA": PT2colors["Not_TCGA"],
                    "ANUS - Not_TCGA": "#EE6C4D", "SI - Not_TCGA": "#FCBF49"}
PTM2colors = {**PT2colors, **PTM2colors_ntcga}

# Primary_Site
x = "Primary_Site"
y = "Project_TCGA_More"
PS_unique = set().union(*[set(df[x]) for df in dfs_cln.values()])
PS2PTM = get_dict_mode_xy(df=pd.concat([df[[x,y]] for df in dfs_cln.values()]), x=x, y=y)
PS2colors = {k: PTM2colors[v] for k,v in PS2PTM.items()}

# Histological_Type
x = "Histological_Type"
HT_unique =set().union(*[set(df["Histological_Type"].dropna()) for df in dfs_cln.values()])
HT2colors = get_labels2colors(labels_unique=HT_unique, sort=True, cmap_name="tab20")
```

Display colors for `Project_TCGA`.

```{r display PT2colors, fig.width=10, fig.height=8, fig.align="center"}
PT2colors <- py$PT2colors
PT2colors <- PT2colors[sort(names(PT2colors))]
dfs_colors <- add_colors_to_table(PT2colors, name="Project_TCGA", dfs=dfs_colors)
rect_plot_colors(line=10, colors=PT2colors)
```

Display colors for `Project_TCGA_More`.

```{r display PTM2colors, fig.width=10, fig.height=8, fig.align="center"}
PTM2colors <- py$PTM2colors
PTM2colors <- PTM2colors[sort(names(PTM2colors))]
dfs_colors <- add_colors_to_table(PTM2colors, name="Project_TCGA_More", dfs=dfs_colors)
rect_plot_colors(line=10, colors=PTM2colors)
```

Display colors for `Primary_Site`.

```{r display PS2colors, fig.width=10, fig.height=10, fig.align="center"}
PS2colors <- py$PS2colors
PS2colors <- PS2colors[sort(names(PS2colors))]
dfs_colors <- add_colors_to_table(PS2colors, name="Primary_Site", dfs=dfs_colors)
rect_plot_colors(line=20, colors=PS2colors, cex=0.5)
```

# Biopsies fields colors

```{python load dfs_bio}
cohorts = ["prism", "met500", "tcga"]
dfs_bio = {cohort: load_bio(study=cohort, mode="in_design") for cohort in cohorts}

# Biopsy_Site colors
BS_unique = set().union(*[set(df["Biopsy_Site"]) for df in dfs_bio.values()])
BS_in_PS = BS_unique.intersection(PS_unique)
BS_not_in_PS = BS_unique.difference(PS_unique)
BS2colors = {**{x: PS2colors[x] for x in BS_in_PS},
              **get_labels2colors(labels_unique=BS_not_in_PS,
                                  colors_avoid=[PS2colors[x] for x in BS_in_PS],
                                  cmap_name="tab20")}
BS2colors["Lymph nodes"] = "#AEC7E8"
```

Display colors for `Biopsy_Site`.

```{r display BS2colors, fig.width=10, fig.height=14, fig.align="center"}
BS2colors <- py$BS2colors
BS2colors <- BS2colors[sort(names(BS2colors))]
dfs_colors <- add_colors_to_table(BS2colors, name="Biopsy_Site", dfs=dfs_colors)
rect_plot_colors(line=25, colors=BS2colors, cex=0.5)
```

Let us save the colors we just defined.

# Signatures colors

Colors for plots related to mutational signatures analyses.

```{r mut change colors deconstructSigs, fig.height=0.5, fig.width=6, fig.align="center"}
# deconstructSigs colors
mut_change2colors <- lapply(list('C>A'="#999999", 
                                 'C>G'="#E69F00", 
                                 'C>T'="#56B4E9", 
                                 'T>A'="#009E73", 
                                 'T>C'="#F0E442", 
                                 'T>G'="#0072B2"),
                            function(x) grDevices::adjustcolor(x, alpha=0.8))

rect_plot_colors(line=1, colors=mut_change2colors)
```


```{r mut change colors maftools, fig.height=0.5, fig.width=6, fig.align="center"}
# maftools colors
mut_change2colors <- lapply(list('C>T'='coral4',
                                 'C>G'='lightcyan4',
                                 'C>A'='cornflowerblue',
                                 'T>A'='lightsalmon1',
                                 'T>C'='forestgreen',
                                 'T>G'='deeppink3'),
                            function(x) grDevices::adjustcolor(x, alpha=0.8))

mut_change2colors <- mut_change2colors[sort(names(mut_change2colors))]
dfs_colors <- add_colors_to_table(mut_change2colors, name="Signatures", dfs=dfs_colors)
rect_plot_colors(line=1, colors=mut_change2colors)
```


```{r variants colors, echo=T, fig.height=1, fig.width=10, fig.align="center", fig.cap="Variants colors"}
#  Variant classification colors
#  Used in 
#  - maftools::plotmafSummary
#  - maftools::oncoplot
VC_ensembl2colors  <- lapply(list(missense_variant = '#33a02c',
                          stop_gained      = '#e31a1c',
                          frameshift_del   = '#1f78b4',
                          frameshift_ins   = '#ff7f00',
                          hotspot          = '#6a3d9a',
                          inframe_ins      = '#a6cee3',
                          inframe_del      = '#fdbf6f',
                          start_lost       = '#b2df8a',
                          stop_lost        = '#cab2d6'),
                     function(x) adjustcolor(x, alpha=0.8))

VC_ensembl2colors <- VC_ensembl2colors[sort(names(VC_ensembl2colors))]
dfs_colors <- add_colors_to_table(VC_ensembl2colors, name="VC_ensembl", dfs=dfs_colors)
rect_plot_colors(line=2, colors=VC_ensembl2colors)
```

# Save in excel workbook

```{r save colors, include=T}
filepath <- file.path(folder_results, "colors_template.xlsx")
wb <- createWorkbook(type="xlsx")

for (name in names(dfs_colors)){
  sheet <- createSheet(wb, sheetName=name)
  setColumnWidth(sheet, colIndex=1, colWidth=35)
  addDataFrame(dfs_colors[[name]], sheet, startRow=1, startColumn=1, row.names=F)
}


saveWorkbook(wb, filepath)
cat(paste("Excel workbook with colors saved at", filepath, "\n"))
```

# Session Info

```{r r config}
sessionInfo()
```
