---
title: "META-PRISM: Analysis of fusions"
author:
  - Daniel Gautheret^[I2BC, CNRS, CEA & Université Paris-Saclay, 91190 Gif-Sur-Yvette, France
      (daniel.gautheret@i2bc.paris-saclay.fr)]
  - Yoann Pradat^[MICS Lab CentraleSupelec, 91190 Gif-Sur-Yvette, France (yoann.pradat@centralesupelec.fr)]
date: "`r format(Sys.time(), '%d %B, %Y')`"
bibliography: ../assets/references.bib
link-citations: true
params:
  cohorts: null
  counts: null
  fusions: null
  samples: null
  tables_burden: null
  tables_driver_fus: null
  tables_driver_gen: null
output:
  bookdown::html_document2:
    theme: cosmo
    css: ../assets/configure.css
    toc: true
    toc_float: true
    code_folding: show
  rmdformats::material:
    highlight: kate
---

<style>
pre {
  overflow-x: auto;
}
pre code {
  word-wrap: normal;
  white-space: pre;
}
body {
  text-align: justify
}
</style>

```{r hidden paramers, include=F}
knitr::opts_knit$set(root.dir = '../..')
library(stringr)
nrows_render_tables <- 5 # set to -1 to render full tables
params_cohorts <- unlist(strsplit(params$cohorts, split="\\s"))
params_fusions <- unlist(strsplit(params$fusions, split="\\s"))
params_samples <- unlist(strsplit(params$samples, split="\\s"))
params_tables_burden <- unlist(strsplit(params$tables_burden, split="\\s"))
params_tables_driver_fus <- unlist(strsplit(params$tables_driver_fus, split="\\s"))
params_tables_driver_gen <- unlist(strsplit(params$tables_driver_gen, split="\\s"))
```

This file is a short report presenting the key steps of the fusions analysis pipeline along with some key figures. The
cohorts analyzed in this pipeline are `r paste(params_cohorts, collapse=", ")`

# Libraries

```{r R setup libraries, message=F}
options(run.main=FALSE)
source("workflow/functions/io.R")
source("workflow/scripts/02.2_plot_burden.R")
source("workflow/scripts/03.2_plot_driver.R")

library(readr)

# in-house packages
library(rprism)
library(rprismtools)
```

# Data used

In this analysis pipeline, data from different sources was used. In particular,

- **curated clinical tables** where used for each cohort. These were produced after careful manual review and automatic
  assembly of multiple annotation files. They are loaded using the `rprism::load_cln` function.

- **aggregated fusion calls** file for each cohort.
  * **MET500** fusion calls are available on the nextcloud at `r rprism:::get_filepath_fusions("met500")`. All
    fusions called by one of the fusion callers described in the fusion pipeline description notebook.
  * **PRISM** fusion calls are available on the nextcloud at `r rprism:::get_filepath_fusions("prism")`. All fusions
    called by one of the fusion callers described in the fusion pipeline description notebook.
  * **TCGA** fusion calls are available on the nextcloud at `r rprism:::get_filepath_fusions("tcga")`. Supplementary
    file of the paper [@gao_driver_2018].


# Preprocessing

## Fusions annotation

In order to align the META-PRISM fusion calling pipeline with that employed on the TCGA data in [@gao_driver_2018],
only fusions called jointly by **Arriba** or **Starfusion** were retained. The details of the rationale for this choice
are detailed in the rna pipeline description file.

Further filtering was applied in order to purify the final list of fusions analyzed. For all fusions of TCGA, PRISM and
MET500, only those fusions that

* are in the COSMIC white list (see `cosmic-V92-fusions-curated.tsv` in the folder
  "resources/fusions_analysis/fusion_lists" on the nextcloud)
* are in the ChiTaRS white list (see `Chitars-TIC.tsv` in the folder
  "resources/fusions_analysis/fusion_lists" on the nextcloud)
* are not in the Bibaceanu black list (see `blacklist-Bibaceanu-enhanced.tsv` in the folder
  "resources/fusions_analysis/fusion_lists" on the nextcloud)

were retained.

Eventually, for the **driver analysis**, only fusions where at least one of the involved genes is known to be a cancer
driver (COSMIC cancer gene census list v92, only Tier 1 genes) or fusions that are listed in the COSMIC database of
fusions were analyzed. 

The script "01.1_preprocess_fusions.R" loads all the fusion calls for each cohort and add filtering annotations for the
filters described above. It also adds some patients and sample annotations that are useful for grouping samples in some
of the anlyses. The tables below show the first lines of annotated fusion call tables for each cohort.

```{r read fusions, include=F}
dfs_fus <- load_fusions_for_analysis(params_cohorts, params_fusions, NULL, NULL)
```

<div style = "width:100%; height:auto; margin: auto;">
```{r show fus prism, echo=F}
render_table(dfs_fus[["prism"]], caption="Annotated fusions PRISM", extensions=c("Responsive"))
```
</div>

<div style = "width:100%; height:auto; margin: auto;">
```{r show fus met500, echo=F}
render_table(dfs_fus[["met500"]], caption="Annotated fusions MET500", extensions=c("Responsive"))
```
</div>

<div style = "width:100%; height:auto; margin: auto;">
```{r show fus tcga, echo=F}
render_table(dfs_fus[["tcga"]], caption="Annotated fusions TCGA", extensions=c("Responsive"))
```
</div>

## Samples selection 

As not all cohorts have samples for all tumor types, the list of tumor types analyzed was restricted. As the focus of
ths study is on samples from the PRISM cohort, the set of tumor type was restricted to tumor types present in PRISM with
a sufficient number of samples. For some analyses, especially for the **driver analysis** where we compare the
prevalence of some events between all cohorts, it was necessary to further restrict the list of tumor types analyzed
to the intersection of tumor types between all cohorts. The table below summarizes the number of samples available for
each tumor type in each cohort and provides binary indicators that indicate whict tumor type was used in which analysis.

```{r load counts tt, include=F}
df_counts_tt <- read_tsv(params$counts, col_types=cols())
```

<div style = "width:100%; height:auto; margin: auto;">
```{r show tumor type counts, echo=F}
render_table(df_counts_tt, caption="Tumor types counts and selection", nrows=-1, extensions=c("Responsive"))
```
</div>

# Fusions burden 

In this analsyis, the fusions burden of each tumor type was compared between all cohorts.  The analysis is performed in
the script **02.1_burden_tables.R** and **02.2_plot_burden.R**. The following figure summarizes this comparison for
common tumor types.

```{r burden plot, echo=F, message=F, fig.width=7.1, fig.height=5, fig.align="center", dpi=140}
# load
dfs_burden <- setNames(lapply(params_tables_burden, read_tsv, col_types=cols()), params_cohorts)
dfs_burden <- lapply(dfs_burden, function(df) df %>% replace(.==0, 0.5))

# plot
cohorts2colors <- load_colors(sheet="Global")[params_cohorts]
draw_burden_plot_fusion(dfs_burden, cohorts2colors)
```


# Driver fusions

In this analsyis, the prevalence of some fusions was compared between all cohorts as well as the prevalence of genes
involved in fusions. For the latter, only "cancer driver genes" were considered (see above). The analysis is performed in
the script **03.1_driver_tables.R** and **03.2_plot_driver.R**. The following figures summarizes both comparisons for
common tumor types.

Here is the figure for fusions.

```{r driver plot fus, echo=F, warning=F, fig.width=6.4, fig.height=3.2, fig.align="center", dpi=300}
dfs_plot <- read_plot_tables(params_tables_driver_fus)
cohorts2colors <- load_colors(sheet="Global")[c("tcga", "met500")]
stacks2colors <- list(`Known COSMIC`="#B3DE6A", Novel="#80B1D3", `Known ChiTaRS only`="#FDB462")
col_var <- "Fusion_Id"
col_stack <- NULL

# plot
fig <- draw_prevalence_plot(dfs_plot, col_var=col_var, cohorts2colors=cohorts2colors,
                            alpha_left=0.05, alpha_heatmap=0.05,
                            name_left_pvals="pvalsstrats", name_heatmap_pvals="pvals",
                            col_stack=col_stack, stacks2colors=stacks2colors)
fig <- fig %>% layout(autosize=F, width=900, height=450, margin=list(b=0))
fig
```

Here is the figure for "cancer driver genes" involved in fusions.

```{r driver plot gen, echo=F, warning=F, out.width="100%"}
dfs_plot <- read_plot_tables(params_tables_driver_gen)
cohorts2colors <- load_colors(sheet="Global")[c("tcga", "met500")]
stacks2colors <- list(`Known COSMIC`="#B3DE6A", Novel="#80B1D3", `Known ChiTaRS only`="#FDB462")
col_var <- "Hugo_Symbol"
col_stack <- "Fusion_Status"

# plot
fig <- draw_prevalence_plot(dfs_plot, col_var=col_var, cohorts2colors=cohorts2colors,
                            alpha_left=0.05, alpha_heatmap=0.05,
                            name_left_pvals="pvalsstrats", name_heatmap_pvals="pvals",
                            col_stack=col_stack, stacks2colors=stacks2colors)
fig <- fig %>% layout(autosize=F, width=1200, height=600, margin=list(b=0))
fig
```


# Notebook configuration

Here is the R configuration

```{r r config}
sessionInfo()
```

`r if (knitr::is_html_output()) '# References {-}'`
